---
title: "Chapter 2 notes"
author: "Addison"
date: "4/17/2022"
output: html_document
---

# Chapter 2 Time series graphics [[textbook link]](https://otexts.com/fpp3/graphics.html)

Quote: "Graphs enable many features of the data to be visualised, including patterns, unusual observations, changes over time, and relationships between variables"

Load required packages.
```{r, library-loading, results='hide', warning=FALSE, message=FALSE}
library(fpp3)
library(readr)
```

## 2.1 `tsibble` objects [[section link]](https://otexts.com/fpp3/tsibbles.html)

Think of a time series as a list of numbers (measurements) along with info about the times they were recorded.

### The index variable [[subsection link]](https://otexts.com/fpp3/tsibbles.html#the-index-variable)

```{r, first-tsibble-example}
# tsibble is used for temporal data
y <- tsibble(
  Year = 2015:2019,
  Observation = c(123, 39, 78, 52, 110),
  index = Year
)
y
```

```{r, monthly-data}
# Some example monthly observational data
mon <- c("2019 Jan", "2019 Feb", "2019 Mar","2019 Apr", "2019 May")
obs <- c(50, 23, 34, 30, 25)
z <- data.frame(Month = mon, Observations = obs)

# Creating a tsibble from a data frame of temporal data
z %>%
  # yearmonth, below, is an example of a time class function
  # other examples in Freq/Func table in section 2.1
  # from annual down to sub-daily.
  mutate(Month = yearmonth(Month)) %>%
  as_tsibble(index=Month)
```

### The key variables [[subsection link]](https://otexts.com/fpp3/tsibbles.html#the-key-variables)

```{r, multiple-time-series-in-single-tsibble}
olympic_running

# Show the categories of a variable
olympic_running %>% distinct(Sex)

# Show the distinct combinations of two variables.
# Section 2.1 states these two series uniquely identify
# each record.
olympic_running %>% distinct(Sex, Length)

```

### Working with `tsibble` objects [[subsection link]](https://otexts.com/fpp3/tsibbles.html#working-with-tsibble-objects)

Working with dataset `PBS`. Medicare Australia prescription data from July 1991 to June 2008. These are classified according to various concession types, and Anatomical Therapeutic Chemical (ATC) indexes.

```{r, common-dplyr-functions}
PBS

# We're interested in the Cost time series
PBS %>% 
  filter(ATC2 == "A10")

# We're interested in a few select columns
PBS %>% 
  filter(ATC2 == "A10") %>% 
  select(Month, Concession, Type, Cost)

# We want to see the total cost per month
# regardless of Concession or Type
PBS %>% 
  filter(ATC2 == "A10") %>% 
  select(Month, Concession, Type, Cost) %>% 
  summarise(TotalC = sum(Cost))

# change the units from dollars to millions of dollars
# and save result to new variable.
PBS %>% 
  filter(ATC2 == "A10") %>% 
  select(Month, Concession, Type, Cost) %>% 
  summarise(TotalC = sum(Cost)) %>% 
  mutate(Cost = TotalC / 1e6) -> a10

a10
```

### Read a csv file and conver to a tsibble [[subsection link]](https://otexts.com/fpp3/tsibbles.html#read-a-csv-file-and-convert-to-a-tsibble)

Commonly one reads in csv data, and then identifies the index and key variables.

```{r, read-in-csv}
prison <- readr::read_csv("remote_data/prison_population.csv")

prison <- prison %>% 
  mutate(Quarter = yearquarter(Date)) %>%
  # select everything except the Date column
  select(-Date) %>% 
  as_tsibble(
    key = c(State, Gender, Legal, Indigenous),
    index = Quarter
  )

prison
```

Book states: "for a tsibble to be valid, it requires a unique index for each combination of keys." Therefore, we get an error with the following:
```{r, unique-index-test}
tryCatch(
  expr = data.frame(
    Year = c(1990,1990,1990,1990),
    key1 = c("3","3","3","3"),
    key2 = c("4","5","5","6"),
    m  = c(1,2,3,4)
  ) %>% as_tsibble(index = Year, key = c(key1, key2)),
  error = function(x) {
    x
  }
)
```

```{r}
print("tryCatch worked as expected: printed error and execution continued.")
```

### The seasonal period [[subsection link]](https://otexts.com/fpp3/tsibbles.html#the-seasonal-period)

Contains a useful table containing numbers of time units in different periods.

## 2.2 Time plots

"For time series data, the obvious graph to start with is a time plot. That is, the observations are plotted against the time of observation, with consecutive observations joined by straight lines."

```{r}
melsyd_economy <- ansett %>% 
  filter(Airports == "MEL-SYD", Class == "Economy") %>% 
  mutate(Passengers = Passengers / 1000)

# Automatically select a plot given the data passed in
autoplot(melsyd_economy, Passengers) +
  labs(title = "Ansett airlines economy class",
       subtitle = "Melbourne-Sydney",
       y = "Passengers ('000)")
```
```{r}
# relies on a10 data from section 2.1
autoplot(a10, Cost) +
  labs(y = "$ (millions)",
       title = "Australian antidiabetic drug sales")
```

## 2.3 Time series patterns

More careful definitions of "trend" and "seasonal" would be useful.

### Trend

Long period of overall increase or decrease. There may be seasonal ups and downs, but overall there is a general direction. Does not have to be linear.

### Seasonal

Seasonal patterns are where there is a clear effect that lines up with some time period like yearly or weekly. Seasonality is always associated with a fixed time period length.

### Cyclic

Like seasonal, but not associated with a necessarily fixed time period length. Often associated with economic events, or what "business cycles." The duration is typically at least 2 years.

Cyclic and seasonal often confused for one another, but in actuality they are quite different. Figure 2.3 shows different combinations of these components. [[Link to figure]](https://otexts.com/fpp3/tspatterns.html#fig:fourexamples).

1. Top left: No apparent trend. Strong yearly seasonal pattern. Apparent 6-10 year cycle.
2. Top right: Clear downward trend. No clear seasonality. Book mentions more data might reveal the trend is part of a cycle. _I think I can see a potential cycle on the order of 25 years._
3. Bottom left: Clear upward trend. Apparent strong seasonality. No evidence of cyclic component.
4. Bottom right: No clear trend, seasonality, or cyclic behavior. Apparently no strong patterns that would help develop a forecasting model.

## 2.4 Seasonal Plots

Like time plot except each season broken out into separate series for the plot.

```{r}
a10 %>%
  gg_season(Cost, labels = "both") +
  labs(y = "$ (millions)",
       title = "Seasonal plot: Antidiabetic drug sales")
```
Aspects of seasonal plots:
- Can see seasonal patterns clearly
- Can see deviations between seasons more easily
- Useful, for example, in seeing years in which patterns change

In the above plot
- you can see a repeating bump in sales recorded in January each year
- you can see dip in March 2008 sales unlike other years
- the dip in June 2008 probably due to sales data not completely collected yet for that month.

### Multiple seasonal periods [[subsection link]](https://otexts.com/fpp3/seasonal-plots.html#multiple-seasonal-periods)

The seasonal plot function, `gg_seasonal`, allows you to plot different period lengths with the `period` argument.

```{r}
vic_elec %>% 
  gg_season(Demand, period = "day") +
  theme(legend.position = "none") +
  labs(y = "MWh", title = "Electricty demand: Victoria")
```
Above, you can see 3 lines with 25 hours, which correspond to uncorrected daylight savings times.

```{r}
vic_elec %>% 
  gg_season(Demand, period = "week") +
  theme(legend.position = "none") +
  labs(y = "MWh", title = "Electricty demand: Victoria")
```
Above you can see weekly seasonal patterns.

```{r}
vic_elec %>% 
  gg_season(Demand, period = "year") +
  theme(legend.position = "none") +
  labs(y = "MWh", title = "Electricty demand: Victoria")
```
Here you can see the yearly seasonal patterns.

## 2.5 Seasonal subseries plots [[subsection link]](https://otexts.com/fpp3/subseries.html#subseries)

You can also create subplots, one per month, with the domain over the different years.

```{r}
a10 %>% 
  gg_subseries(Cost) +
  labs(y = '$ (millions)',
       title = 'Australian antidiabetic drug sales')
```
The book mentions the blue line indicating the seasonality, but I don't find it particularly easy to conclude that from looking at this plot. The book goes on to say the plot isn't particularly revealing, with which I agree. I am curious what it would look like if there was a subplot per year, and the domain going over that year.

I tried to change the domain with the following, but it didn't do what I was aiming for.

```{r}
# This does not do the thing. Anyone know how?

# a10 %>% 
#   gg_subseries(Cost) +
#   labs(y = '$ (millions)',
#        title = 'Australian antidiabetic drug sales',
#        period = 'year')
```
```{r}
head(tourism)

holidays <- tourism %>% 
  filter(Purpose == 'Holiday') %>% 
  group_by(State) %>% 
  summarise(Trips = sum(Trips))

head(holidays)
```

```{r}
autoplot(holidays, Trips) +
  labs(y = "Overnight trips ('000)",
       title = 'Australian domestic holidays')

```
```{r}
gg_season(holidays, Trips) +
  labs(y = "Overnight trups('000)",
       title = "Australian domestic holidays")
```
```{r}
holidays %>% 
  gg_subseries(Trips) +
  labs(y = "Overnight trips ('000)",
       title = "Australian domestic holdays")
```
Above, you can see most of the regions have seen an increase over the mean in the past few years.

## Scatterplots [[section link]](https://otexts.com/fpp3/scatterplots.html)

```{r}
vic_elec %>%
  filter(year(Time) == 2014) %>%
  autoplot(Demand) +
  labs(y = "GW",
       title = "Half-hourly electricity demand: Victoria")
```
```{r}
vic_elec %>%
  filter(year(Time) == 2014) %>%
  autoplot(Temperature) +
  labs(
    y = "Degrees Celsius",
    title = "Half-hourly temperatures: Melbourne, Australia"
  )
```

```{r}
vic_elec %>%
  filter(year(Time) == 2014) %>%
  ggplot(aes(x = Temperature, y = Demand)) +
  geom_point() +
  labs(x = "Temperature (degrees Celsius)",
       y = "Electricity demand (GW)")

```

Figures 2.14 and 2.15 display correlations between various variables. Wikipedia article on Correlation: [[wikipedia]](https://en.wikipedia.org/wiki/Correlation).

### Scatterplot matrices [[subsection link]](https://otexts.com/fpp3/scatterplots.html#scatterplot-matrices)

```{r}
visitors <- tourism %>%
  group_by(State) %>%
  summarise(Trips = sum(Trips))
head(visitors)
```

```{r}
visitors %>%
  ggplot(aes(x = Quarter, y = Trips)) +
  geom_line() +
  facet_grid(vars(State), scales = "free_y") +
  labs(title = "Australian domestic tourism",
       y= "Overnight trips ('000)")
```
```{r, results='hide', warning=FALSE, message=FALSE}
library(GGally)
```

```{r}
visitors %>%
  pivot_wider(values_from=Trips, names_from=State) %>%
  GGally::ggpairs(columns = 2:9)
```




